Grow with the flow: a latitudinal cline in growth rate is associated with more variable precipitation in \textit{Mimulus cardinals}

### 2016-10-12 ms/ms.Rnw
Worked on analyses, results, tables

%* show figures of reaction norms
%* split plot, account for error variation in temp???
%* show PAR similarity
%* what CO2 concentration did I use?
%* some sites they are functionally annual

@@

%* check out matthew's blog on hydrology variables

@@

%* put in biomass allocation cline
% TO DO:
% 1. Is water regime on figure 1 correct?
% 2. Finalize Table 2
% 3. Do anything with biomass data?
% 4. Citations for latitudinal clines
	% cases where clinal variation has 'obvious' selective source:
	% Antonovics and Bradshaw discuss steepness of clines and strength of selection over 10's of meters off mines in Anthoxanthum
	% Caicedo et al 2004?
	% clines in phenotypic traits reviewed by Hedrick 2006, Schemske and Bier 2007
% 5. Citations for GxE as signature of local adaptation
% 6. Fix figure 2
% 7. Supplemental Figure and Table captions
% 8. Supplemental figure showing agreement between assigned to climatic variables from point estimated and spatially averaged analyses
% 9. new spatially averaged analysis
% 10. new analysis with updated ENM
% 11. mention using LER_AbsGrowth as predictor in fitSER
% 12. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)

@@

  # Small function to make statistical significance asterisks
  sigStar <- function(pvalue)
  {
    if (pvalue >= 0.05) return("")
    if (pvalue < 0.05 & pvalue >= 0.01) return("*")
    if (pvalue < 0.01 & pvalue >= 0.001) return("**")
    if (pvalue < 0.001) return("***")
  }

@@

\title{Grow with the flow: faster growth is associated with more variable precipitation in a perennial herb}
\author{Christopher D. Muir$^{1,*}$ and Amy L. Angert$^1$}
\date{}

@@

Original Article \\

@@

Word counts: TBA \\
3 figures; 3 tables; ? supporting files \\
Data will be archived on Dryad upon acceptance. \\

\cdm{Chris' comments}\\
\ala{Amy's comments}

@@

\subsection*{Treatments}

@@

\begin{table}[ht]
   \centering
   \topcaption{Key traits measured in this study.}
   \begin{tabular}{@{} ll @{}}
      \toprule
  Trait & Units \\
      \midrule
  Day of germination  & day \\
  Leaf expansion rate  &  mm day$^{-1}$  \\
  Shoot elongation rate  &  mm or cm day$^{-1}$  \\
  Harvest dry mass &  g  \\
  Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
  Mortality &  \cdm{Probability?}  \\
	    \bottomrule
   \end{tabular}
   \label{table:Table_traits}
\end{table}

\paragraph{Day of germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}. The model was fit with Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function. The signifcance of the Population effect was determined using analysis of deviance.

@@

We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements. We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.  We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing). Both leaf expansion and shoot elongation were modeled as a second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.

@@

\paragraph{Photosynthesis}
During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of \Sexpr{length(which(!is.na(master$Photo)))} plants evenly spread between treatments and families within populations. The youngest, fully-expanded leaf \cdm{this is what I did...not sure exactly which node, as I didn't record that when I did measurements} acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska). All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity. During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions.

@@

\paragraph{Mortality}
We assayed mortality during twice-weekly growth measurements. We analyzed the probability of surviving until the end of the experiment as a function of population, treatment, and their interactions using a Generalized Linear Mixed Model (GLMM) assuming binomially distributed errors. As before, we included Family and Block as random factors. We assessed significance of fixed effects using Type 2 Analysis of Deviance with Wald \Chi$^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.

@@

	# Code for GLM (remove if satisfied with GLMER)
  # fitMort <- glm(DiedFromStress ~ PopID * TempTrt * WaterTrt,
  #  data = subset(master, !is.na(master$DiedFromStress)), 
  #  family = "quasibinomial")

@@

\cdm{show correlation between growth rate and biomass?}

@@

For all traits (Table~\ref{table:Table_traits}) we tested for Population, Treatment, and Population $\times$ Treatment interactions. We interpreted significant Population effects to indicate intrinsic variation and Population by Treatment effects to indicate variation in plasticity. As mentioned above, we survival and GLM models for germination rate and mortality, respectively. For all other traits, we used mixed model ANOVAs with Family included as a random factor. Models were fit by restricted maximum likelihood using lmer from the R package \pkg{lme4} \citep{Bates_etal_2014}. Significant fixed effect terms were selected using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-11 \citep{Kuznetsova_etal_2014}. Denominator degrees of freedom for $F$-tests were estimated using Satterthwaite's approximation. Significant Population effect indicate intrinsic trait differences; significant Population $\times$ Treatment effects indicate population differences in plasticity. For growth rate, we also accounted for differences in germination rate by including day of germination as a factor.

\subsection*{Principal components of germination, growth, and photosynthesis}
For each single-trait model above, we extracted the Population coefficient (factoring out Treatment and other effects). The multivariate distribution of these coefficients was then summarized using principal components analysis (PCA). The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast and slow growing populations.

@@

We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We looked for overlap between climatic variables that best predict latitude of \textit{M. cardinalis} occurrence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations. For brevity, we refer to these as Climate-Latitude and Climate-Trait variables. We selected Climate-Latitude and Climate-Trait variables independently using Random Forest (VSURF algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}). From VSURF models, we kept only variables selected for prediction, the most stringent criterion.

If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-Trait analyses \ala{and the correlation magnitudes should be high? maybe this language is not technically correct (variable importance more accurate?) but the point is that this prediction needs something about count and the strength}. If latitude is correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline, then several climatic variables should be weakly correlated \ala{fix this to mirror prior}.

To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-Trait variables as above, except that we used spatially averaged climate variables. \ala{what's the actual prediction? that spatially averaged Climate-Trait will have higher correlation than point estimated Climate-Trait?} We sampled climate at 1000 random points (at 90-m resolition) within a 100-km buffer around focal populations. We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}. Since \textit{M. cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}. Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}. For clarity, we distinguish between analyses where climate is inferred from a single point (`point estimated Climate-Trait') versus averaged across a 100-km buffer (`spatially averaged Climate-Trait').

@@

	# Calculatins for plot dimensions
	sw1a <- strwidth(s = clim2sym[imp1, "climFactor"], units = "inches")
	sw2a <- strwidth(s = clim2sym[imp1, "tempScale"], units = "inches")
	sw3a <- strwidth(s = parse(text = clim2sym[imp1, "what"]), units = "inches")

	sw1b <- strwidth(s = clim2sym[imp2, "climFactor"], units = "inches")
	sw2b <- strwidth(s = clim2sym[imp2, "tempScale"], units = "inches")
	sw3b <- strwidth(s = parse(text = clim2sym[imp2, "what"]), units = "inches")

@@

	text(grconvertX(panelGap / 2, from = "in", to = "user"), 
	     nImpLat:1, pos = 4, labels = clim2sym[imp1, "climFactor"])
  text(grconvertX(panelGap + max(c(sw1a, sw1b)), from = "in", to = "user"), 
       nImpLat:1, pos = 4, labels = clim2sym[imp1, "tempScale"])
	text(grconvertX(3 / 2 * panelGap + max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)), 
	                               from = "in", to = "user"), nImpLat:1, pos = 4,
	                labels = parse(text = clim2sym[imp1, "what"]))

@@

	text(grconvertX(yLabWidth + 2 * panelWidth + 3 / 2 * panelGap, 
	                from = "in", to = "user"), 
	     nImpLat:(nImpLat - nImpPC1 + 1), pos = 4, 
	     labels = clim2sym[imp2, "climFactor"])
  text(grconvertX(yLabWidth + 2 * panelWidth + 5 / 2 * panelGap + 
                    max(c(sw1a, sw1b)), from = "in", to = "user"), 
       nImpLat:(nImpLat - nImpPC1 + 1), pos = 4, 
       labels = clim2sym[imp2, "tempScale"])
	text(grconvertX(yLabWidth + 2 * panelWidth + 5 / 2 * panelGap + 
                    max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)), from = "in", 
	                to = "user"), nImpLat:(nImpLat - nImpPC1 + 1), pos = 4,
	                labels = parse(text = clim2sym[imp2, "what"]))

@@

For these analyses, we compiled a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M. cardinalis} occurences. These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included \Sexpr{length(climVars) / 2} climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}). The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range. For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation. Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.

@@

Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M. cardinalis}, as evidenced by large and highly significant population effects (Table~\ref{table:Table_AnovaSummary}). A single principal component captured \Sexpr{gsub("%", "", pve)}\% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}). As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table of Population, Treatment, Population x Treatment interactions for varies traits
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

   \topcaption[ANOVA summary]{Summary of Population, Treatment, and Population $\times$ Treatment effects. We used different statistical modeling for the diverse traits assayed -- glmer: generalized linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}. Note that temperature and water treatments were imposed after germination, hence are not application to this trait. Key to statistical significance: *$P < 0.05$; ** $P < 0.01$; *** $P < 0.001$}

@@

\begin{figure}[h!]
	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude. Each point is a population, its latitude of origin a position along the slow to fast growth axis, defined as Principal Component 1 of five traits (see Methods). The line and 95\% confidence intervals were estimated using standard linear regression.}
	\label{fig:Fig_PC1vLat}
\end{figure}

@@

Genotype $\times$ environment interactions are also a common signature of local adaptation. For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions. We found remarkably little evidence for this pattern in \textit{M. cardinalis}. There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing. Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}). Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects. However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.

@@

% ANOVA table for LER (need to label tables)
\begin{table}[htbp]
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[ANOVA table, leaf expansion rate]{CAPTION}
	\begin{center}
	\begin{tabular}{lcccccc}
	\toprule

	\input{Tables/Table_lllGrowthAnova.txt}

>>>>

%* show figures of reaction norms
%* split plot, account for error variation in temp???
%* show PAR similarity
%* some sites they are functionally annual

@@

@@

%* put in biomass allocation cline
% TO DO:
% 1. Is water regime on figure 1 correct? (data is in office)

% 3. Do anything with biomass data?
% 4. Citations for latitudinal clines
	% cases where clinal variation has 'obvious' selective source:
	% Antonovics and Bradshaw discuss steepness of clines and strength of selection over 10's of meters off mines in Anthoxanthum
	% Caicedo et al 2004?
	% clines in phenotypic traits reviewed by Hedrick 2006, Schemske and Bier 2007
% 5. Citations for GxE as signature of local adaptation
% 7. Supplemental Figure and Table captions
% 8. Supplemental figure showing comparison between assigned to climatic variables from point estimated and spatially averaged analyses
% 9. new spatially averaged analysis
% 12. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)

@@

  # Source custom functions
  source(paste(pathMS, "/functions.R", sep = ""))

@@

\title{Grow with the flow: a latitudinal cline in growth rate is associated with more variable precipitation in \textit{Mimulus cardinals}}
\author{Christopher D. Muir$^{1,*}$ and Amy L. Angert$^1$}
\date{}

@@

Running Head: Latitudinal cline and climate in \textit{Mimulus} \\

@@

Data will be archived on Dryad upon publication. \\

@@

\subsection*{Temperature and drought treatments}

@@

\begin{table}[ht]
   \centering
   \topcaption{Key traits measured in this study.}
   \begin{tabular}{@{} ll @{}}
      \toprule
  Trait & Units \\
      \midrule
  Day of germination  & day \\
  Leaf expansion rate  &  mm day$^{-1}$  \\
  Shoot elongation rate  &  mm or cm day$^{-1}$  \\
  Harvest dry mass &  g  \\
  Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
  Mortality & probability of death  \\
	    \bottomrule
   \end{tabular}
   \label{table:Table_traits}
\end{table}

\paragraph{Day of germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}. The model was fit with Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function. The signifcance of the Population effect was determined using analysis of deviance. Note that, unlike other traits discussed below, we did not include Block or Population $\times$ Treatment interactions because during germination plants had not been placed into blocks and treatments had not been applied.

@@

We measured growth rate in two phases: leaf expansion as a rosette and shoot elongation after bolting. We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements. We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.  We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing). Both leaf expansion and shoot elongation were modeled as second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.

@@

\paragraph{Photosynthesis}
During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of \Sexpr{length(which(!is.na(master$Photo)))} plants evenly spread between treatments and families within populations. The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska). All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity. During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions. As expected, stomatal conductance was strongly correlated with photosynthetic rate (results not shown), but we found that treatment and population effects were more apparent when stomatal conductance was factored out. Therefore, we calculated the `intrinsic' photosynthetic rate for each individual as its residual deviance from the regression between stomoatal conductance and photosynthetic rate across all plants. We also removed one outlier.

@@

\paragraph{Mortality}
We assayed mortality during twice-weekly growth measurements. We analyzed the probability of surviving until the end of the experiment as a function of population, treatment, and their interactions using a Generalized Linear Mixed Model (GLMM) assuming binomially distributed errors. We included Family and Block as random effects. We assessed significance of fixed effects using Type 2 Analysis of Deviance with Wald \Chi$^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.

@@

@@

We harvested plants that survived until the end of the experiment from 3 to 16 July (80 to 93 days after sowing). We separated individual plants into three biomass fractions: leaves, shoots, and belowground (roots plus rhizomes). Belowground fractions were rinsed thoroughly in water to remove sand. After completely drying in a forced air oven (Shel Lab, Sheldon Manufacturing Inc., Oregon, USA), we weighed the dry mass of each fraction using an analytical balance.

@@

For all traits (Table~\ref{table:Table_traits}) except germination (see above), we tested for Population, Treatment, and Population $\times$ Treatment interactions. We interpreted significant Population effects to indicate intrinsic variation and Population $\times$ Treatment interactions to indicate variation in plasticity. As mentioned above, we used survival and GLMM models for germination rate and mortality, respectively. For all other traits, we used mixed model ANOVAs with Family and Block included as random factors. Models were fit by restricted maximum likelihood using lmer from the R package \pkg{lme4} \citep{Bates_etal_2014}. Significant fixed effect terms were selected using a step-wise backward elimination procedure implemented with the step function in the R package \pkg{lmerTest} version 2.0-11 \citep{Kuznetsova_etal_2014}. Denominator degrees of freedom for $F$-tests were estimated using Satterthwaite's approximation. To ensure that Population and Treatment effects were specific to a particular growth phase, we included earlier growth phases as a covariate. Specifically, we included germination day as a covariate in leaf expansion and shoot elongation analyses; we included germination day and leaf expansion rate as a covariate in shoot elongation analyses.

\subsection*{Principal components of germination, growth, and photosynthesis}
For each single-trait model above, we extracted the Population coefficient (factoring out Treatment and other effects). The multivariate distribution of these coefficients was then summarized using principal components analysis (PCA). The first principal component of these traits (Trait PC1) loaded positively with germination, growth, and photostynthetic rate, therefore we define this as a phenotypic axis dilineating fast to slow growth.

@@

We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We looked for overlap between climatic variables that best predict latitude of \textit{M. cardinalis} occurrence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations. For brevity, we refer to these as `Climate-Latitude' and `Climate-TraitPC1' variables. We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}. From VSURF models, we kept only variables selected for prediction, the most stringent criterion.

If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-TraitPC1 analyses \ala{and the correlation magnitudes should be high? maybe this language is not technically correct (variable importance more accurate?) but the point is that this prediction needs something about count and the strength}. If latitude is correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline, then several climatic variables should be weakly correlated \ala{fix this to mirror prior}.

To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-TraitPC1 variables as above, except that we used spatially averaged climate variables. \ala{what's the actual prediction? that spatially averaged Climate-TraitPC1 will have higher correlation than point estimated Climate-Trait?} We sampled climate at 1000 random points (at 90-m resolition) within a 62-km buffer around focal populations. We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}. Significant spatial autocorrelation persisted for approximately 62 km. Since \textit{M. cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}. Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}. We distinguish between analyses where climate is inferred from a single point (`point estimated Climate-TraitPC1') versus averaged across a 62-km buffer (`spatially averaged Climate-TraitPC1').

@@

	# Calculatins for plot dimensions
	sw1a <- strwidth(s = clim2sym[imp1, "climFactor"], units = "inches", font = 4)
	sw2a <- strwidth(s = clim2sym[imp1, "tempScale"], units = "inches", font = 4)
	sw3a <- strwidth(s = parse(text = clim2sym[imp1, "what"]), units = "inches",
	                 font = 4)

	sw1b <- strwidth(s = clim2sym[imp2, "climFactor"], units = "inches", font = 4)
	sw2b <- strwidth(s = clim2sym[imp2, "tempScale"], units = "inches", font = 4)
	sw3b <- strwidth(s = parse(text = clim2sym[imp2, "what"]), units = "inches",
	                 font = 4)

@@

	text(grconvertX(panelGap / 2, from = "in", to = "user"), 
	     nImpLat:1, pos = 4, labels = clim2sym[imp1, "climFactor"],
	     font = ifelse(imp1 %in% intersect(imp1, imp2), 2, 1))
  text(grconvertX(panelGap + max(c(sw1a, sw1b)), from = "in", to = "user"), 
       nImpLat:1, pos = 4, labels = clim2sym[imp1, "tempScale"],
       font = ifelse(imp1 %in% intersect(imp1, imp2), 2, 1))
	text(grconvertX(3 / 2 * panelGap + max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)), 
	                               from = "in", to = "user"), nImpLat:1, pos = 4, 
	     labels = parse(text = clim2sym[imp1, "what"]),
	     font = ifelse(imp1 %in% intersect(imp1, imp2), 2, 1))

@@

	text(grconvertX(yLabWidth + 2 * panelWidth + 3 / 2 * panelGap, 
	                from = "in", to = "user"), 
	     nImpLat:(nImpLat - nImpPC1 + 1), pos = 4, 
	     labels = clim2sym[imp2, "climFactor"],
	     font = ifelse(imp2 %in% intersect(imp1, imp2), 2, 1))
  text(grconvertX(yLabWidth + 2 * panelWidth + 5 / 2 * panelGap + 
                    max(c(sw1a, sw1b)), from = "in", to = "user"), 
       nImpLat:(nImpLat - nImpPC1 + 1), pos = 4, 
       labels = clim2sym[imp2, "tempScale"],
       font = ifelse(imp2 %in% intersect(imp1, imp2), 2, 1))
	text(grconvertX(yLabWidth + 2 * panelWidth + 5 / 2 * panelGap + 
                    max(c(sw1a, sw1b)) + max(c(sw2a, sw2b)), from = "in", 
	                to = "user"), nImpLat:(nImpLat - nImpPC1 + 1), 
	     pos = 4, labels = parse(text = clim2sym[imp2, "what"]),
	     font = ifelse(imp2 %in% intersect(imp1, imp2), 2, 1))

@@

For these analyses, we compiled a representative set of \Sexpr{nrow(thinnedOccClimate)} recent (since 2000) known \textit{M. cardinalis} occurences. These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA version 5.30 \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included \Sexpr{length(climVars) / 2} climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}). The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range. For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation. Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.

@@

Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M. cardinalis}, as evidenced by large and significant population effects (Table~\ref{table:Table_AnovaSummary}). A single principal component captured \Sexpr{gsub("%", "", pve)}\% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}). As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted with the low amount of variation in plasticity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table of Population, Treatment, Population x Treatment interactions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

   \topcaption[ANOVA summary]{Summary of Population, Treatment, and Population $\times$ Treatment effects. We used different statistical modeling for the diverse traits assayed -- glmer: generalized linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}. Note that temperature and water treatments were imposed after germination, hence are not applicable to this trait. Complete analysis of variance/deviance tables for each trait are available in the Supporting Information. Key to statistical significance: *$P < 0.05$; ** $P < 0.01$; *** $P < 0.001$}

@@

\begin{figure}[h!]
	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude. Each point is a population's latitude of origin (x-axis) and position along the slow to fast growth axis (y-axis), defined as Principal Component 1 of five traits (see Methods). The line and 95\% confidence intervals were estimated using linear regression.}
	\label{fig:Fig_PC1vLat}
\end{figure}

@@

Genotype $\times$ environment (G$\times$E) interactions are also a common signature of local adaptation. For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions. We found little evidence G$\times$E in \textit{M. cardinalis}. \cdm{Is this result still correct after revising analysis?} There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing. Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}). Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects. However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.

@@

% ANOVA table for LER (need to label tables)

<<TableS_fitLER, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=
	  source(paste(pathMS, "/functions.R", sep = ""))

  ss <- sigDig(fitLER$anova.table$`Sum Sq`)
  ms <- sigDig(fitLER$anova.table$`Mean Sq`)
  df1 <- sigDig(fitLER$anova.table$NumDF)
  df2 <- sigDig(fitLER$anova.table$DenDF)
  fval <- sigDig(fitLER$anova.table$F.value)
 pval <- pval2latex(fitLER$anova.table$`Pr(>F)`)
@

\begin{table}[htbp]
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[ANOVA table, leaf expansion rate]{CAPTION}
	\begin{center}
	\begin{tabular}{lcccccc}
	\toprule

                     & SS & MS & df1 & df2 & \em{F} & $P$ \\ \midrule
  Day of Germination & 18.2 & 18.2 & 1 & 633.1 & 35.2 & 4.78 $\times10^{-9}$ \\
mmLER

Population & 22.3 & 1.5 & 15 & 121.5 & 4.3 & 2.57 $\times10^{-6}$ \\
Temperature & 114.4 & 114.4 & 1 & 567.7 & 352.3 & 1.64 $\times10^{-61}$ \\
Water & 4.2 & 4.2 & 1 & 567.7 & 12.6 & 4.12 $\times10^{-4}$ \\
Population $\times$ Temperature & 3.1 & 0.2 & 15 & 553.3 & 0.6 & 0.86 \\
Population $\times$ Water & 8.9 & 0.6 & 15 & 567.8 & 1.8 & 0.03 \\
Temperature $\times$ Water & 0 & 0 & 1 & 550.5 & 0 & 0.99 \\
Population $\times$ Temperature $\times$ Water & 4 & 0.3 & 15 & 535.4 & 0.8 & 0.69 \\

	\bottomrule
	\end{tabular}
	\label{table:TableS_fitLER}
	\end{center}
\end{table}
